---
title: "Reproducible Research: Peer Assignment 1"
author: "Bill Killacky"
date: "January 18, 2014"
output: html_document
---

```{r computetime,echo=FALSE}
    time <- format(Sys.time(), "%a %b %d %Y %X")
    # http://astrostatistics.psu.edu/su07/R/html/base/html/strptime.html 
    # %a weekday abbr
    # %b month abbr
    # %d dayOfMonth 1-31
    # %X Time local 
    # %Y YYYY
    # setwd("/Users/billkillacky/myRstuff/ReproducableResearch")
    # same as setwd("~/myRstuff/ReproducableResearch")
  # cloned version
```

Report creation time: `r time`.


```{r,echo=TRUE}
    setwd("C:/Users/Bill Killacky/Documents/Coursera/ReproducableResearch/RepData_PeerAssessment1/RepData_PeerAssessment1")
    w <- getwd()
    f <- 'activity.csv'
    f2 <- paste(w, '/activity.zip',sep='')
    unzip(f2)

    library(plyr)
    library(xtable)
    library(ggplot2)
    
    # warn if input file is not in working directory    
    if (!file.exists(f)) {
        print(paste('input file',f,'is not in working directory:',w))
    }
```

  
##Loading and preprocessing the data
* Grading: Does the submission show code for reading in the dataset and/or processing the data? 

```{r readfile,results='asis'}
    # 1.    Load the data 
    a <- read.csv(file=f, stringsAsFactors=FALSE)
    
    # 2.    Process/transform the data into a steps per day analysis format
    s <- ddply(a, .(date), summarize,
                 TotSteps=sum(steps), 
                 meanSteps=mean(steps), 
                 medianSteps=median(steps)
               )
```

##What is mean total number of steps taken per day?
* Grading: Are both the mean and median number of steps taken each day reported?  

```{r meanmedian, results='asis'}
    
    # Calculate and report the mean and median total number of steps taken per day
    xt <- xtable(s)
    print(xt, type="html")  

```


* Grading: Does the submission contain a histogram of the total number of steps taken each day?  

```{r hist}
    # 1.    Make a histogram of the total number of steps taken each day
    hist(s$TotSteps, breaks=26, col='cyan', ylim=c(0,20))
```

##What is the average daily activity pattern?
* Grading: Is there a time series plot of the average number of steps taken (averaged across all days) versus the 5 minute intervals?

```{r plot}
    is <- aggregate(a$steps, by=list(interval=a$interval), mean, na.rm=TRUE)
    colnames(is) <- c('interval', 'meanSteps')
    plot(is$interval, is$meanSteps, type='l', col='blue', ylim=c(0, 240))
```  
  

###Most Active 5-minute interval  
* Grading: Does the report give the 5-minute interval that, on average, contains the maximum number of steps?  
Which 5-minute interval contains the max number of steps?  

```{r maxint,results='asis'}
    xt <- is[is$meanSteps==max(is$meanSteps),]
    xt <- xtable(xt)
    print(xt, type="html") 
```  


##Imputing Missing Values
1.    Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)  


```{r naDiff}
    nf <- nrow(a)                     # number of rows in original dataset
    nna <- nrow(a[is.na(a$steps),])   # number of rows with steps = NA
    nnaPct <- nna/nf                  # percent of original dataset with NA
    nok <- nf - nna                   # number of rows that do not have NA steps
    
    fn <- sprintf('%d', nf)
    fnna <- sprintf('%d',nna)
    fnnaPct <- paste(sprintf('%d', floor(nnaPct*100)), '%')
    fok <- sprintf('%d', nok)
    
    Description <- c('number of rows with steps = NA', 'number of rows in original dataset', 
                'number of rows that do not have NA steps', 'percent of original dataset with NA')
    Rows <- c(fnna, fn, fok, fnnaPct)
```

Missing Value Statistics  
```{r naDiffShow,results='asis'}
    xt <- data.frame(Description, Rows)
    xt <- xtable(xt)
    print(xt, type="html")   
  
```


###Impute Plan: Replace NA with the mean steps for that 5-minute interval
2.  Strategy for filling in all of the missing values in the dataset:  **Use the mean for that 5-minute interval**.
3.  Create a new dataset (m) that is equal to the original dataset (a) but with the missing data filled in.  

* Grading: Does the report describe and show with code a strategy for imputing missing data?  


```{r naHist}
    # Create new dataset m that is equal to the original dataset but with the missing data filled
    # in with the mean steps matching on the 5-minute interval
    #
    m <- merge(is, a, by.x='interval', by.y='interval')
    n <- nrow(m)
    for (i in 1:n) {
        if (is.na(m$steps[i])) {
            m$steps[i] <- m$meanSteps[i]     # replace missing steps data with the interval mean
        }
    }

head(a[a$date=='2012-10-01',])    # data snapshot Before Impute Plan
head(m[m$date=='2012-10-01', ])   # new data snapshot After Impute Plan (Note steps NA replaced with interval mean value)
```

* Grading: Does the report contain a histogram of the total number of steps taken each day after missing values were imputed?

```{r imputedfile}
    
    # 2.    Process/transform the new dataset without NAs into a steps per day analysis format
    sm <- ddply(m, .(date), summarize,
                 TotSteps=sum(steps), 
                 meanSteps=mean(steps), 
                 medianSteps=median(steps)
               )

    # 1.    Make a histogram of the total number of steps taken each day
    hist(sm$TotSteps, breaks=26, col='cyan', ylim=c(0,20), main='Histogram of Daily Total Steps Taken AFTER Imputing')
    hist(s$TotSteps, breaks=26, col='gray', ylim=c(0,20), main='(Compare with histogram above)\nHistogram of Daily Total Steps Taken BEFORE Imputing')
s[is.na(s$TotSteps),]  # Show Dates formerly without data
sm[is.na(s$TotSteps),] # Show Dates with Imputed Data

```

####4. Impact of imputing missing data
4a.  * Make a histogram of the total number of steps taken each day and  
**Answer: Please refer to color histogram above ("AFTER Imputing")**  

4b.  * Calculate and report the mean and median total number of steps taken per day.  
**Answer: Please refer to table below.**  

4c.  * Do these values differ from the estimates from the first part of the assignment?   
**Answer: While the average per interval remains the same, these new values include an additional daily total steps of 10,766 each for the eight days formerly without data**  

4d.  * What is the impact of imputing missing data on the estimates of the total daily number of steps?    
**Answer: The addition of eight days of 10,766 total steps also adds 8 to the frequency in the center of the histogram**  
 

```{r NoNAmeanmedian, results='asis'}

    # Report the mean and median total number of steps taken per day
    library(xtable)
    xt <- xtable(sm)
    print(xt, type="html")  

```

  
####Updated Missing Value Statistics  


```{r updatedIntervalPlot}
    # Recalculate steps by interval using new dataset with replaced means for the old NAs.
    #
    is2 <- aggregate(m$steps, by=list(interval=m$interval), mean, na.rm=TRUE)
    colnames(is2) <- c('interval', 'meanSteps')
    #plot(is2$interval, is2$meanSteps, type='l', col='green', ylim=c(0, 240))
```


```{r naDiff2,results='asis'}
    #
    # After Imputing Missing Values, verify that NAs are gone and total rows matches original dataset.
    #
    nf <- nrow(m)                     # number of rows in original dataset
    nna <- nrow(m[is.na(m$steps),])   # number of rows with steps = NA
    nnaPct <- nna/nf                  # percent of original dataset with NA
    nok <- nf - nna                   # number of rows that do not have NA steps
    
    fn <- sprintf('%d', nf)
    fnna <- sprintf('%d',nna)
    fnnaPct <- paste(sprintf('%d', floor(nnaPct*100)), '%')
    fok <- sprintf('%d', nok)
    
    Description <- c('number of rows with steps = NA', 'number of rows in original dataset', 
                'number of rows that do not have NA steps', 'percent of original dataset with NA')
    Rows <- c(fnna, fn, fok, fnnaPct)

    xt <- data.frame(Description, Rows)
    xt <- xtable(xt)
    print(xt, type="html")   
  
```


##Are There Differences in Activity Patterns between Weekdays and Weekends?

```{r PlotInOne}
    #
    # Determine if a date is weekend or a weekday
    #
    m$date2 <- as.POSIXlt(m$date)
    m$wday <- m$date2$wday
    m$weekdays <- weekdays(m$date2)
    m$dayType <- ifelse(m$wday %in% c(0,6), 'weekend', 'weekday')   #0=Sun, 1=M, 2=Tu, 3=W, 4=Th, 5=F, 6=Sat
    #head(m, 7)
    # What are the counts of the different types of day (weekend, weekday)
    table(m$dayType)

```



```{r}
    mm <- ddply(m, .(interval, dayType), summarize, 
                intSumSteps=sum(steps), 
                intMeanSteps=mean(steps))
    #head(mm,6)
    g <- ggplot(mm, aes(interval, intMeanSteps))
    g <- g + geom_line(aes(color=dayType))
    g <- g + geom_smooth(size=4, linetype=3, method='loess', se=FALSE)
    g <- g + facet_grid(dayType~.)   # multiple panels result from facet
    g <- g + labs(title = 'Activity: Weekday vs Weekend') + labs(x='Intervals', y='Interval Mean Steps')
    g
   
```

* Grading: Does the report contain a panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends?

###Weekends and Weekday Patterns are Different
* Note: the square blue boxes are smoothing of the interval activity through out the day.
* Overall weekends tend to be more active than weekdays (based on the number of blue boxes above 50)
* Weekends start later and end later with more activity in the middle of the day.
* Weekdays start sooner with a spike of activity and are relatively inactive until the less active evening commute home.
* Weekdays also cease activity earlier perhaps indicating an earlier bed time during the work week.

* Grading: Does the report show all of the R code needed to reproduce the results (numbers, plots, etc.) in the report?

